Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-04 revised_main.tex
Updates from Overleaf

% need to simplify the intro, and expand the hypothesis
\section{Introduction}
Since the 1970s, artists such as Suzanne Ciani~\cite{ciani_life_in_waves} have explored the use of audio synthesizers to replicate natural sounds—such as a bottle cap popping off—in ways that preserve the essential qualities of the original while showcasing the unique characteristics of the synthesizer~\cite{creativecherep2024}. Rather than merely replicating sounds, this approach uses synthesizers for artistic reinterpretation. However, the automation of this interpretive approach to sound design using digital audio synthesizers has rarely been explored, and significant knowledge gaps remain even in the simpler problem of replicating sounds. One such gap is the lack of diversity and depth in the analysis of sound-similarity measures and their interaction with various synthesis methods. As a stepping stone toward automated interpretive sound design, we use automatic and manual sound-evaluations to investigate whether the optimal choice of a sound-similarity measure is dependent on the synthesis method.

A digital audio synthesizer is any software used for the creation and manipulation of audio. The possibilities afforded by these synthesizers is infinite, leading to their widespread adaption by artists and sound-designers~\cite{lyons1997understanding,russ1999sound,stranneby2004digital}. A typical synthesizer has a number of parameterizable functions which affect the output sound in various ways, and the manual approach involves modifying the parameters until a desired sound is reached.

The automation of the aforementioned approach is often referred to as "sound-matching", with reduction of tinkering time and creation of ``interesting" sounds as the main motivators \cite{krekovic2019insights,turian2020sorry,horner1993machine,salimi2020make,esling2019flow,engel2020ddsp,mitchell2007evolutionary,shier2020spiegelib,masuda2021soundmatch,masuda2023improving}. Sound-matching involves a target sound and a similarity metric (or loss function), and a heuristic to find parameters for a synthesizer that replicate \textit{all} or \textit{some} of the characteristics of a target sound as best as possible~\cite{horner1993machine,mitchell2007evolutionary,masuda2023improving}.

Through our review of the history of sound-matching, we provide a nomenclature for its different subtypes, and illuminate major perennial issues that prevent sound-matching as a practical approach for sound-designers. Here we are concerned with the issues of \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the---perhaps needless---focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods. As a consequence of these problems, not much is known about the interaction between different methods of synthesis and different loss functions. While there have been many works and experiments comparing the accuracy of loss functions and choosing the best performers~\cite{vahidi2023mesostructures,turian2020sorry,engel2020ddsp,uzrad2024diffmoog,han2023perceptual,masuda2021soundmatch,turian2020sorry,bruford2024synthesizer}, we note that these claims regarding the effectiveness of one function versus another have consistently been made in limited contexts that may not generalize to other settings as typically only a single (and often, simple) synthesizer has been used for testing the effectiveness of loss functions. Should we be searching for a universally best performing loss function across synthesizers? Does the method of synthesis influence the selection of the ``optimal" loss function, and if so, do current, SOTA loss functions satisfy the creative needs of sound-designers?

The main hypothesis here is that the performance of a similarity measure (or loss function) is influenced by other factors in the environment, particularly the method of synthesis. For the experiments, a \textit{differentiable}, \textit{iterative} approach to sound-matching is used. The iterative approach  better mimics the manual process of sound design (recursively listening to synthesizer outputs and adjusting the parameters accordingly) while allowing access to the gradients that can be used to better understand the nature of the problem.

>>>>

% need to simplify the intro, and expand the hypothesis
\section{Introduction}
Since the 1970s, artists such as Suzanne Ciani~\cite{ciani_life_in_waves} have explored the use of audio synthesizers to replicate natural sounds—such as a bottle cap popping off—in ways that preserve the essential qualities of the original while showcasing the unique characteristics of the synthesizer~\cite{creativecherep2024}. Rather than merely replicating sounds, this approach uses synthesizers for artistic reinterpretation. However, the automation of this interpretive approach to sound design using digital audio synthesizers remains impractical for sound-designers, and significant knowledge gaps remain even in the simpler problem of replicating sounds. One such gap is the lack of diversity and depth in the analysis of sound-similarity measures and their interaction with various synthesis methods. As a stepping stone toward automated interpretive sound design, we use automatic and manual sound-evaluations to investigate whether the optimal choice of a sound-similarity measure is dependent on the synthesis method.

A digital audio synthesizer is any software used for the creation and manipulation of audio. The possibilities afforded by these synthesizers is infinite, leading to their widespread adaption by artists and sound-designers~\cite{lyons1997understanding,russ1999sound,stranneby2004digital}. A typical synthesizer has a number of parameterizable functions which affect the output sound in various ways, and the manual approach often involves modifying the parameters until a conceptualized target sound is reached.

The automation of the aforementioned approach has been commonly referred to as "sound-matching", with reduction of tinkering time and creation of ``interesting" sounds as the main motivators \cite{krekovic2019insights,turian2020sorry,horner1993machine,salimi2020make,esling2019flow,engel2020ddsp,mitchell2007evolutionary,shier2020spiegelib,masuda2021soundmatch,masuda2023improving}. Sound-matching involves a target sound and a similarity metric (or loss function), and a heuristic to find parameters for a synthesizer that replicate \textit{all} or \textit{some} of the characteristics of a target sound as best as possible~\cite{horner1993machine,mitchell2007evolutionary,masuda2023improving}.

We provide a review of past works in order to understand what is lacking in the field, and how it can be improved. Through our review of the history of sound-matching, we provide a nomenclature for its different subtypes, and illuminate major perennial issues that prevent sound-matching as a practical approach for sound-designers. Here we are concerned with the issues of \LossSelect~and \SynthSelect. The former refers to the search for an optimal loss function, lack of novel algorithms, and the---perhaps needless---focus on outperforming state of the art (\gls{SOTA}), while the latter refers to the lack of diversity in synthesis methods. As a consequence of these problems, we find that not much is known about the interaction between different methods of synthesis and different loss functions. While there have been many works and experiments comparing the accuracy of loss functions and choosing the best performers~\cite{vahidi2023mesostructures,turian2020sorry,engel2020ddsp,uzrad2024diffmoog,han2023perceptual,masuda2021soundmatch,turian2020sorry,bruford2024synthesizer}, we note that these claims regarding the effectiveness of one function versus another have consistently been made in limited contexts that may not generalize to other settings as typically only a single (and often, simple) synthesizer has been used for testing the effectiveness of loss functions. Should we be searching for a universally best performing loss function across synthesizers? Does the method of synthesis influence the selection of the ``optimal" loss function, and if so, do current, SOTA loss functions satisfy the creative needs of sound-designers?

The main hypothesis here is that \textit{the performance of a similarity measure (or loss function) is influenced by other factors in the environment, particularly the method of synthesis}. For the experiments, a \textit{differentiable}, \textit{iterative} approach to sound-matching is used. The iterative approach  better mimics the manual process of sound design (recursively listening to synthesizer outputs and adjusting the parameters accordingly) while allowing access to the gradients that can be used to better understand the nature of the problem.
